## Framingham Case Study

The Framingham Heart Study led to the publication of more than 3,000 journal articles, the most highly cited of which is Gordon et al. *Am J Med* 1977, a seminal paper that linked high HDL cholesterol with lower risk for CHD.

We will write a literate program that uses data from the Framingham study to examine whether we can reproduce the link between HDL cholesterol and outcomes such as CHD or death from any cause.

### Install the tidyverse packages, if necessary

```{r, results='hide'}
if (!"tidyverse" %in% installed.packages()) install.packages("tidyverse")
```

### Download the Framingham data

Download `framingham.xlsx` from the course website. 

```{r, results='hide'}
download.file("https://skadauke.github.io/cdar_course/framingham.xlsx",
              destfile = "framingham.xlsx")
```

### Load the data

First load the `tidyverse` and `readxl` packages for data manipulation and functions and to load Excel files.

Then load the file `framingham.xlsx` into a data frame named `framingham`.

```{r, results="hide", message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)

framingham <- read_excel("framingham.xlsx")
```

### Look at the data

> Examine the structure of the data using the `str()` function.

```{r, echo=F, results='hide'}
str(framingham)
```

> Use the `summary()` function on `framingham` to create a brief statistical summary of the distribution of data in each of its columns.

```{r, echo=FALSE, results='hide'}
summary(framingham)
```

> What proportion of participants were smokers at the time of enrollment? What proportion had CHD?
> What proportion of participants developed CHD at the end of the follow up period? What proportion of participants died?

*Hint*: You don't need to enter any additional R code to answer these questions. Take a look at the output of the `summary()` command and remember that if the values of a vector are only `0` and `1` ( or `TRUE` and `FALSE`), the mean of the vector is equal to the proportion of values that are `1` (or `TRUE`).

From the above look at the data, it appears that a number of rows are missing HDL cholesterol data. To be exact, there are a total of 236 rows with missing HDL data. 

> What is the proportion of rows with missing HDL data?

```{r, echo=FALSE, results='hide'}
mean(is.na(framingham$hdl))
```

