## Framingham Case Study

The Framingham Heart Study led to the publication of more than 3,000 journal articles, the most highly cited of which is Gordon et al. *Am J Med* 1977, a seminal paper that linked high HDL cholesterol to lower risk for CHD.

We will be using a data set made available by the National Heart, Lung, and Blood Institute (NHLBI) that summarizes clinic visit data (history, physical, labs) as well as long-term outcomes for participants of the study. The data can be requested [here](https://biolincc.nhlbi.nih.gov/teaching/). A **code book** that explains the study design as well as each variable and its values can be found [here](framingham_code_book.pdf).

For the sake of simplicity, we will use a truncated data set containing only the data from one examination period and a reduced number of columns. The variable (column) names have also been edited for clarity.

We will write a literate program that uses data from the Framingham study to examine whether we can reproduce the link between HDL cholesterol and outcomes such as CHD or death from any cause.

Please note that this data has been treated with statistical methods designed to protect patient confidentiality while largely (but not fully) preserving the structure of the data. Analyses derived from this data are therefore unsuitable for publication.

### Install necessary packages

When R Markdown "knits" the HTML document the R commands in all code chunks get executed. For example, the following code chunk loads all the packages required for this document to work and installs any missing packages. It is a good idea to include code that makes sure all dependencies are met so that others (or future-you) can reproduce the analysis. 

You will probably also want to run this code **now** to make sure you have all the packages installed to run all of the code in the remaining code chunks. To run a code chunk click the little green arrow on the top right of the code chunk in the RStudio editor. Do it now!

```{r, results='hide', message=FALSE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(Hmisc, tidyverse, downloader, readxl, ggplot2)
detach("package:Hmisc", unload=TRUE) # prevent Hmisc from overwriting summarize()
```

### Download the Framingham data

Download `framingham.xlsx` from the course website. (Run the code chunk!)

```{r}
download("https://skadauke.github.io/cdar_course/framingham.xlsx",
        destfile = "framingham.xlsx")
```

### Load the data

Load the file into a data frame named `framingham`. If you run the code chunk below by clicking on the green arrow, `framingham` will be available in your environment. This makes it possible to "play" with the data as you are developing your literate program.

```{r}
framingham <- read_excel("framingham.xlsx")
```

### Look at the data

As you "get to know" a new data set, it is a good idea to jot down observations about the data in the text of a literate program. So let's look at the data and take some notes.

First let's examine the structure of the `framingham` data using the `str()` function. Write your code inside the code chunk below. Remember that you can run the code chunk by clicking the little green arrow on the top right of the code chunk.

```{r}
# YOUR CODE HERE

```

Look at the output above and consider the following questions:

> How many rows does the table have? How many columns?

> Do the column names make sense? 

> How is "any CHD" defined? (Where could you look this up?)

> Do you think each row corresponds to a patient in the study? (Where could you look this up?)

> Are there any columns with missing values?

According to the **code book** for this data set, `sex` is coded as follows:

```
1 = male
2 = female
```

It is a good idea to make the data as self-explanatory as possible. We will therefore recode the `sex` column as a factor variable (R's representation of an unordered categorical variable) with the two levels `male` and `female`. Run the code chunk.

```{r}
framingham <- framingham %>%
  mutate(sex = recode_factor(sex,
                      `1` = "male",
                      `2` = "female"))

str(framingham)
```

Next, let's use the `summary()` function on `framingham` to create a brief statistical summary of the distribution of data in each of its columns. Write your code below and run the code chunk.

```{r}
# YOUR CODE HERE

```

Look at the output of `summary()` and consider the following questions:

> Are men and women equally represented in the study?

> What proportion of participants were smokers at the time of enrollment? 

> What proportion of participants had CHD at enrollment?

> What proportion of participants had developed CHD at the end of the follow up period?

> What proportion of participants died?

> What proportion of participants were *at risk* for developing CHD?

*Hint*: You don't need to enter any additional R code to answer these questions. Take a look at the output of the `summary()` command and remember that if the values of a vector are only `0` and `1` ( or `TRUE` and `FALSE`), the mean of the vector is equal to the proportion of values that are `1` (or `TRUE`).

From the above look at the data, it appears that a number of rows are missing HDL cholesterol data. To be exact, there are a total of 236 rows with missing HDL data. 

Let's figure out what the proportion of rows is that have missing data in the `hdl` column of the `framingham` data frame. Write your code below and run the code chunk.

```{r}
# YOUR CODE HERE

```

Before we continue, let's remove any rows that don't have a value for `hdl` and store the filtered data set in the variable `hdl_measured`. Fill in the blanks below and run the code.

```{r}
# FILL IN THE BLANKS
hdl_measured <- framingham %>%
  ___(___(___))
```

### Examine the relationship between HDL and all-cause mortality

Let's see whether there might be a dose-dependent effect of HDL on all-cause mortality. We will use the powerful **split-apply-combine** conceptual framework to think through this. 

First we need to **split** the data into several chunks. Since we're interested in outcomes at different HDL levels, we split the data by HDL level. Say we want to split at HDLs of 25, 35, 45, 55, 65, and 75.

The first step will be to create a vector `cut_points` with those numbers. Use the `seq()` function to create a vector with the above numbers and enter the code in the chunk below. (Hint: `?seq` !) Don't forget to run the code chunk.

```{r}
# FILL IN THE BLANKS
cut_points <- seq(___, ___, ___)
```

Let's walk through the following pipeline to take a close look of how split-apply-combine works.

We start with the `hdl_measured` data frame, which contains all of the data from `framingham` except rows without HDL measurements.

The `mutate()` function adds a column `hdl_level` which identifies the HDL level "bucket" each row is in. Since there are 6 cut points, there are 7 buckets.

The `group_by()` function does the actual **splitting**, by the `hdl_level` column we just generated. 

`summarize()` then **applies** the `mean()` function to `outcome_death_any_cause` which calculates the proportion of death outcomes at each `hdl_level`. Afterwards, it **combines** the all the calculated proportions into a new data frame.

Note that the resulting data frame has a completely different structure than the data frame `hdl_measured` we fed into the pipeline. This is typical of split-apply-combine workflows.

```{r}
hdl_measured %>%
  mutate(hdl_level = Hmisc::cut2(hdl, cut_points)) %>%
  group_by(hdl_level) %>%
  summarize(death = mean(outcome_death_any_cause))
```

> Modify the above code to break the data down by `sex` in addition to `hdl_level` and rerun the chunk.

> Does it look like HDL is a good predictor of all-cause mortality?

### Examine the relationship between HDL and incidence of CHD

We'd like to know if a low HDL value can predict the development of CHD. Some of the patients already had CHD at baseline so we already know that they will have CHD at the end. 

To find the population **at risk** for developing CHD let's remove all patients that have CHD at baseline. This is captured in the variable `history_chd`. Fill in the blanks in the code below and run the code chunk.

```{r}
# FILL IN THE BLANKS
at_risk <- hdl_measured %>%
  ___(___)
```

Let's do a split-apply-combine operation similar to the one in the previous section.

1. Create a new column `hdl_level` which contains the results of the `Hmisc::cut2` function already filled in.
2. Break down the data frame by `hdl_level` and `sex`.
3. Create a summary table with the following two columns:
    a. `n`: number of patients in the particular group.
    b. `outcome_chd`: proportion of patients that developed CHD (this is coded as `outcome_any_chd`).

```{r}
outcome_chd_by_hdl_level <- at_risk %>%
  ___(___ = Hmisc::cut2(hdl, cut_points)) %>%
  ___(___, ___) %>%
  ___(___ = ___,
      ___ = ___)

outcome_chd_by_hdl_level
```

Let's plot the data! (Unfortunately we didn't get to covering plots in this preview!)

```{r}
ggplot(outcome_chd_by_hdl_level, 
       aes(x = hdl_level, 
           y = outcome_chd, 
           color = sex, 
           size = n)) + 
  geom_point()
```

### Next steps...

- Logistic regression?
- Predict outcomes?
- ...